---
title: "HW 3"
author: "Iris Zhong"
date: '2022-10-25'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nlme)
```


```{r}
data <- read_csv("processed_data.csv")
```

Aggregate by week per person

```{r}
data_by_month <- data %>% 
  group_by(user_id, month = month(start_time)) %>% 
  summarize(month_avg = mean(correct),
            assignment_count_per_month = uniqueN(assignment_id),
            problem_count_per_month = uniqueN(problem_id))
```

```{r}
data_by_month %>%
  group_by(user_id) %>%
  summarize(n = n()) %>%
  ggplot(., aes(x = n)) +
  geom_histogram()
```

```{r}
data_by_month %>%
  group_by(user_id) %>%
  summarize(n_item = sum(problem_count_per_month)) %>%
  ggplot(., aes(x = n_item)) +
  geom_histogram(binwidth = 1) +
  xlim(0, 100)
```


keep persons that have responses in at least 6 months

```{r}
data_by_month <- data_by_month %>%
  group_by(user_id) %>%
  mutate(n_month = n()) %>%
  filter(n_month > 6) %>%
  ungroup()
```

keep persons that have tried at least 10 items each month they participated

```{r}
data_by_month <- data_by_month %>%
  mutate(item_count_per_month_over_10 = ifelse(problem_count_per_month >= 10, 1, 0)) %>%
  group_by(user_id) %>%
  mutate(item_count_per_month_over_10 = mean(item_count_per_month_over_10)) %>%
  filter(item_count_per_month_over_10 == 1) %>%
  select(-item_count_per_month_over_10) %>%
  ungroup()
```



```{r}
data_by_month %>% summarize_all(n_distinct)
```

```{r}
data_by_month <- data_by_month %>%
  select(-c(n_month)) %>%
  group_by(user_id) %>%
  as.tibble() %>%
  complete(user_id, month, 
           fill = list(month_avg = NA, assignment_count_per_month = 0, problem_count_per_month = 0)) %>% 
  as.data.frame()
```

Add number of items/assignments by learner

```{r}
data_by_month <- data_by_month %>%
  group_by(user_id) %>%
  mutate(n_problem = sum(problem_count_per_month),
         n_assignment = sum(assignment_count_per_month))
```

Recode month so that the month number reflects the number of months passed since September. (September coded as 1, October as 2, ..., August as 12)

```{r}
data_by_month <- data_by_month %>%
  mutate(month = case_when(month > 8 ~ month - 8,
                           month <= 8 ~ month + 4))
```


Plot trajectories:

```{r}
ggplot(data = data_by_month,                    #data set
       aes(x = month, y = month_avg, group = user_id)) + #setting variables
  #geom_point(size=.5, alpha = 0.01) + #adding points to plot
  geom_line(alpha = 0.05) +  #adding lines to plot
  theme_bw() +   #changing style/background
  scale_x_continuous(breaks = seq(0, 11)) +
  labs(x = "Month",
       y = "Average score per month",
       title = "Learner average score by month")
```

Centering month (in September: month = 0)

```{r}
data_by_month <- data_by_month %>%
  mutate(month = month - 1)
```

# Model fitting

```{r}
#fitting no growth model and assigning it to an object
mod0 <- nlme(month_avg ~ beta_1 + d_1i,    #model equation
                     data=data_by_month,     #data set                   
                     fixed=beta_1~1,          #fixed parameters              
                     random=d_1i~1,           #random coefficients
                     group=~user_id,               #clustering variable         
                     start=c(beta_1=0.5),      #starting values
                     na.action = na.pass)  #missing data treatment                     

#obtaining summary of the model using the object we just created                     
summary(mod0)
```



```{r}
#obtaining predicted scores for individuals
data_by_month$pred_ng <- predict(mod0)

#obtaining predicted scores for prototype
data_by_month$proto_ng <- predict(mod0, level=0)

#plotting predicted trajectories
#intraindividual change trajetories
ggplot(data = data_by_month, aes(x = month, y = pred_ng, group = user_id)) +
  #geom_point(color="black", alpha = 0.05) + 
  geom_line(color="black", alpha = 0.05) +
  geom_line(aes(x = month, y = proto_ng), color="red",size=2) + 
  theme_bw() +   #changing style/background
  #setting the x-axis with breaks and labels
  scale_x_continuous(breaks = seq(0, 11)) +
  labs(x = "Month",
       y = "Predicted score per month",
       title = "No growth model prediction")
```

```{r}
intervals(mod0)
```



```{r}
#fitting linear growth model and assigning it to an object
mod1 <- nlme(month_avg~(beta_1+d_1i)+(beta_2+d_2i)*month,  
                   data=data_by_month,                      
                   fixed=beta_1+beta_2~1,                      
                   random=d_1i+d_2i~1,
                   group=~user_id,                     
                   start=c(beta_1=0.7,beta_2=0),
                   na.action = na.pass)

#obtaining summary of the model using the object we just created
summary(mod1)
```

```{r}
intervals(mod1)
```



```{r}
#obtaining predicted scores for individuals
data_by_month$pred_lg <- predict(mod1)

#obtaining predicted scores for prototype
data_by_month$proto_lg <- predict(mod1, level=0)

#plotting predicted trajectories
#intraindividual change trajetories
ggplot(data = data_by_month, aes(x = month, y = pred_lg, group = user_id)) +
  #geom_point(color="black") + 
  geom_line(color="black", alpha = 0.05) +
  geom_line(aes(x = month, y = proto_lg), color="red",size=2) + 
  theme_bw() +   #changing style/background
  #setting the x-axis with breaks and labels
  scale_x_continuous(breaks = seq(0, 11)) +
  labs(x = "Month",
       y = "Predicted score per month",
       title = "Linear growth model prediction")
```


Add time-invariant covariates

```{r}
#fitting linear growth model with tic and assigning it to an object
mod2 <- nlme(month_avg ~ (beta_01 + beta_11*n_problem + d_1i) +
                                (beta_02 + beta_12*n_problem + d_2i)*month,
                          data= data_by_month,
                          fixed=beta_01+beta_11+beta_02+beta_12~1,
                          random=d_1i+d_2i~1,
                          group=~user_id,
                          start=c(beta_01=0.7, beta_11=0, 
                                  beta_02=0, beta_12=0),
                          na.action=na.pass)

#obtaining summary of the model using the object we just created
summary(mod2)
```

```{r}
intervals(mod2)
```



```{r}
#obtaining predicted scores for individuals
data_by_month$pred_lg_cov <- predict(mod2)

#obtaining predicted scores for prototype
data_by_month$proto_lg_cov <- predict(mod2, level=0)

#intraindividual change trajetories
#plotting predicted trajectory
ggplot(data = data_by_month, aes(x = month, y = pred_lg_cov, group = user_id)) +
  #geom_point(color="gray", size=.5) + 
  geom_line(alpha = 0.05) +  #adding lines to plot
  theme_bw() +   #changing style/background
  #setting the x-axis with breaks and labels
  scale_x_continuous(breaks = seq(0, 11)) +
  labs(x = "Month",
       y = "Predicted score per month",
       title = "Linear growth model with covariate prediction")
```

